---
title: "1_task"
output: html_document
---
**Task 1: Objective vs Subjective Treatment Response**

**Task 1 Goal and Strategy:**
The treatment may reduce Monthly Migraine Days (MMDs), but this does not always translate into a proportional improvement in patient-reported outcomes (e.g., HIT-6, MIDAS, HADS).
In this task we quantify the objective-subjective mismatch during long-term anti-CGRP treatment.

We proceed in two steps:
**1. Average longitudinal change:** 
We fit separate linear mixed-effects models for MMDs (objective) and HIT-6 (subjective), accounting for repeated measures with a patient-specific random intercept.

**2. Mismatch at Cycle 1:**
We define objective and subjective responders using pre-specified clinical threshold and estimate the proportion of discordant patietns (objective responder , but not subjective responder).

A sensitivity analysis repeats the same workflow under an alternative longitudinal imputation strategy (LOCF) to assess robustness.

```{r setup, message=FALSE, warning=FALSE}
# ---- Setup ----
if (!requireNamespace("rstudioapi", quietly = TRUE)) install.packages("rstudioapi")
library(tidyverse)

# Force working directory to the project root 
setwd(rstudioapi::getActiveProject())
cat("WD set to:", getwd(), "\n")

# Load baseline (clean + imputed version saved in preprocessing step)
baseline_imp <- readRDS("data/processed/baseline_imputed.rds")

# Load analysis datasets created in 02.preprocessing_missing_imputation.Rmd 
data_main <- readRDS("data/processed/data_main_analysis.rds")
data_sens <- readRDS("data/processed/data_sens_analysis.rds")

# Output folders
dir.create("outputs/task1", recursive = TRUE, showWarnings = FALSE)
dir.create("outputs/task1/figures", recursive = TRUE, showWarnings = FALSE)

# Quick checks
cat("Rows (main):", nrow(data_main), "\n")
cat("Row (sens):", nrow(data_sens), "\n")
cat("Unique subjects (main):", n_distinct(data_main$SUBJECT_ID), "\n")
cat("Unique subjects (sens):", n_distinct(data_sens$SUBJECT_ID), "\n")


```
The following setup loads the preprocessed datasets created in the previous steps and defines the analysis environment for Task 1.
Two datasets are used:
- a main dataset based on linear interpolation;
- a sensitivity dataset based on LOCF imputation.
===============================================================================
Analytical dataset: we build a minimal longitudinal dataset for Task 1 including subject  identifiers, time variables (cycle, month), and the two main outcomes (MMDs and HIT-6). Optional variables (HADS-A/D, INT, MIDAS) are retained when available. This dataset will be used both for mixed models and for responder/mismatch definitions.

```{r}
#We require SUBJECT_ID, CYCLE, MONTH, MMDs, HIT6
required_cols <- c("SUBJECT_ID", "CYCLE", "MONTH", "MMDs", "HIT6")
stopifnot(all(required_cols %in% names(data_main)))
stopifnot(all(required_cols %in% names(data_sens)))

keep_cols <- c("SUBJECT_ID", "CYCLE", "MONTH", "MMDs", "HIT6", "HADSA",
               "HADSD","INT", "MIDAS")

# MAIN dataset for Task 1 (interpolation-based)
data_task1_main <- data_main %>%
  # Keep only relevant columns (optional variables are kept only if present)
  dplyr::select(dplyr::any_of(keep_cols)) %>%
  dplyr::mutate(
    SUBJECT_ID = factor(SUBJECT_ID),
    CYCLE = factor(CYCLE),
    # Keep month as an ordered factor (1, 3, 6, 9, 12)
    MONTH = factor(as.character(MONTH),
                   levels = c("1", "3", "6", "9", "12"),
                   ordered = TRUE)
  )

# SENSITIVITY dataset for Task 1 (LOCF-based)
data_task1_sens <- data_sens %>%
  # Keep only relevant columns (optional variables are kept only if present)
  dplyr::select(dplyr::any_of(keep_cols)) %>%
  dplyr::mutate(
    SUBJECT_ID = factor(SUBJECT_ID),
    CYCLE = factor(CYCLE),
    # Keep month as an ordered factor (1, 3, 6, 9, 12)
    MONTH = factor(as.character(MONTH),
                   levels = c("1", "3", "6", "9", "12"),
                   ordered = TRUE)
  )


cat("Task1 MAIN dim:", dim(data_task1_main), "\n")
cat("Missingness MAIN (MMDs):", mean(is.na(data_task1_main$MMDs)), "\n")
cat("Missingness MAIN (HIT6):", mean(is.na(data_task1_main$HIT6)), "\n")

cat("Task1 SENS dim:", dim(data_task1_sens), "\n")
cat("Missingness SENS (MMDs):", mean(is.na(data_task1_sens$MMDs)), "\n")
cat("Missingness SENS (HIT6):", mean(is.na(data_task1_sens$HIT6)), "\n")
```
We created two parallel longitudinal dataset (MAIN and SENS) to run the same analyses under alternative missing-data handling strategies. MONTH is modeled as an ordered factor because follow-up is scheduled at 1, 3, 6, 9 and 12 months. The missingness summaries serve as a sanity check before model fitting.
===============================================================================

Mixed models (average effects over time): We fit two separate linear mixed models with a random intercept for each patient. This estimates the average change over time while accounting for repeated measures within subjects.

```{r task1-mixed-models-main, message=FALSE, warning=FALSE}

# Load required packages
if (!requireNamespace("lme4", quietly = TRUE)) install.packages("lme4")
if (!requireNamespace("lmerTest", quietly = TRUE)) install.packages("lmerTest")
             
library(lme4)
library(lmerTest)

# MAIN dataset only (interpolation-based)
dat <- data_task1_main

# ---- Model A: Objective outcome (Monthly Migraine Days) ----
mod_mmds <- lmer(
  MMDs ~ CYCLE + MONTH + (1 | SUBJECT_ID),
  data = dat,
  REML = FALSE
)

# ---- Model B: Subjective outcome (HIT-6) ----
mod_hit6 <- lmer(
  HIT6 ~ CYCLE + MONTH + (1 | SUBJECT_ID),
  data = dat,
  REML = FALSE
)

summary(mod_mmds)
summary(mod_hit6)

# Optional models (only if the variables exist)
mods_optional <- list()

if("HADSA" %in% names(dat)) {
  mods_optional$hadsa_main <- lmer(HADSA ~CYCLE + MONTH + (1 | SUBJECT_ID), data = dat, REML = FALSE)
}


if("HADSD" %in% names(dat)) {
  mods_optional$hadsd_main <- lmer(HADSD ~ CYCLE + MONTH + (1 | SUBJECT_ID), data = dat, REML = FALSE)
}

if (length(mods_optional) > 0) lapply(mods_optional, summary)

```
MONTH is treated as an ordered factor (1, 3, 6, 9, 12), so R reporst polynomial contrats for time: MONTH.L (overall linear trend), MONTH.Q(curvature), MONTH.C (cubic) and MONTH^4(higher-order).
These mixed models estimate the average change over follow-up while accounting for repeated measurements via a subject-specific random intercept.
Optioanl HADS models are fitted only where available; therefore they involve a smaller sample (only subjects with HADS observed at some visits), so their estimates are not direclty comparable in precision with the MMDs/HIT-6 models.


Export model results: we exported fixed effects tables (estimate + 95% CI) and create mean +SE trend plots.
```{r task1-fixed-effects-main, message=FALSE, warning=FALSE}

# Packages for tidying model outputs
if (!requireNamespace("broom.mixed", quietly = TRUE)) install.packages("broom.mixed")
library(broom.mixed)

# ---- Export fixed effects tables (beta, SE, CI) ----
fx_mmds_main <- tidy(mod_mmds, effects = "fixed", conf.int = TRUE)
fx_hit6_main <- tidy(mod_hit6, effects = "fixed", conf.int = TRUE)

dir.create("outputs/task1", recursive = TRUE, showWarnings = FALSE)

readr::write_csv(fx_mmds_main, "outputs/task1/fixed_effects_mmds_main.csv")
readr::write_csv(fx_hit6_main, "outputs/task1/fixed_effects_hit6_main.csv")

# Optional models (if you fitted them in mods_optional)
if (exists("mods_optional") && length(mods_optional) > 0) {
  fx_optional_main <- purrr::imap_dfr(
    mods_optional,
    ~ tidy(.x, effects = "fixed", conf.int = TRUE) %>% dplyr::mutate(model = .y)
  )
  readr::write_csv(fx_optional_main, "outputs/task1/fixed_effects_optional_main.csv")
}

# Print in console for quick inspection
fx_mmds_main
fx_hit6_main
```
Fixed-effects estimates (β), standard errors and 95% confidence intervals were exported for the MMDs and HIT-6 mixed models. These tables summarize the average association of cycle and follow-up month via each outcome, while accounting for within-patient repeated measurements via a random intercept. Optional models (HADS-A/D) were exported only when available.

===============================================================================
**Core plots (MAIN):** We summarise MMDs and HIT-6 at each scheduled visit (months 1, 3, 6, 9 ,12) within each treatmentcycle and visualize mean ± standard error. This provides an interpretable view of average longitudinal trends and acts as a sanity check before the mismatch/responder analysis.
```{r task1-core-plots-main, message=FALSE, warning=FALSE}

# ---- Trend plots: mean ± SE by (CYCLE, MONTH) ----

#Core plots (MAIN dataset): mean ± trends by (CYCLE, MONTH)
dir.create("outputs/task1/figures", recursive = TRUE, showWarnings = FALSE)

# Compute summary statistics per timepoint
trend_main <- data_task1_main %>%
  group_by(CYCLE, MONTH) %>%
  summarise(
    n_MMDs = sum(!is.na(MMDs)),
    mean_MMDs = mean(MMDs, na.rm = TRUE),
    se_MMDs = sd(MMDs, na.rm = TRUE) / sqrt(n_MMDs),

    n_HIT6 = sum(!is.na(HIT6)),
    mean_HIT6 = mean(HIT6, na.rm = TRUE),
    se_HIT6 = sd(HIT6, na.rm = TRUE) / sqrt(n_HIT6),
    .groups = "drop"
  )

# Plot: MMDs over time
p_mmds_main <- ggplot(trend_main, aes(x = MONTH, y = mean_MMDs, group = CYCLE, 
                                 color = CYCLE)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_MMDs - se_MMDs, ymax = mean_MMDs + se_MMDs), width = 0.1) +
  labs(
    title = "MMDs over time (MAIN)",
    x = "Month",
    y = "Mean MMDs"
  )

# Plot: HIT-6 over time
p_hit6_main <- ggplot(trend_main, aes(x = MONTH, y = mean_HIT6, group = CYCLE, 
                                      color = CYCLE)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_HIT6 - se_HIT6, ymax = mean_HIT6 + se_HIT6), width = 0.1) +
  labs(
    title = "HIT-6 over time by cycle",
    x = "Month",
    y = "Mean HIT-6"
  )

# Show plots
p_mmds_main
p_hit6_main

# Save plots
ggsave("outputs/task1_trend_mmds_main.png", p_mmds_main, width = 7, height = 4)
ggsave("outputs/task1_trend_hit6_main.png", p_hit6_main, width = 7, height = 4)

# Save trend table too 
readr::write_csv(trend_main, "outputs/task1_trend_summary_main.csv")
```
The resulting figures show the average trajectoreis of objective (MMDs) and subjective (HIT-6) outcomes voer follow-up. We saved both the plots and the aggregated summary table to be reused in the final report.

```{r task1-spaghetti-mmds-main, message=FALSE, warning=FALSE}
set.seed(123)

# Sample a subset of patients to avoid overplotting
sample_ids <- sample(unique(data_task1_main$SUBJECT_ID), 30)

spaghetti_data <- data_task1_main %>%
  filter(SUBJECT_ID %in% sample_ids)

ggplot(spaghetti_data,
       aes(x = MONTH, y = MMDs, group = SUBJECT_ID)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Individual MMDs trajectories (subset of patients)",
    x = "Month within cycle",
    y = "MMDs"
  )
```
As an exploratory visualization, we report individual longitudinal trajectories (spaghetti plot) for a subset of patients, highighting the  substantial inter-individual heterogeneity underlying the average trends.
===============================================================================
Now we move from average longitudinal trends to a patient-level definition of response in Cycle 1.
We construct a start/end dataset by selecting Cycle 1 only and extracting, for each subject, baseline values at Month 1 and end-of-cycle values at Month 12 (or, if Month 12 is missing, the last observed month within Cycle 1, consistent with the "last available visit" rule).

We then compute two change metrics:
- pct_change_MMDs = (MMDs_end − MMDs_start) / MMDs_start (objective percent change; negative values indicate improvement)
- delta_HIT6 = HIT6_end − HIT6_start (subjective absolute change; negative values indicate improvement)

```{r}
# ------------------------------------------------------------------
# Select Cycle 1 and prepare time variable
# ------------------------------------------------------------------

c1 <- data_task1_main %>%
  filter(CYCLE == "1") %>%
  mutate(
    # Convert ordered MONTH factor to numeric for comparisons
    month_num = as.numeric(as.character(MONTH))
  ) %>%
  arrange(SUBJECT_ID, month_num)

cat("Cycle 1 rows:", nrow(c1), "\n")
cat("Unique patients in Cycle 1:", n_distinct(c1$SUBJECT_ID), "\n")


# ------------------------------------------------------------------
# Define start of cycle (Month 1)
# ------------------------------------------------------------------
# Start-of-cycle values are taken at MONTH == 1

c1_start <- c1 %>%
  filter(month_num == 1) %>%
  select(
    SUBJECT_ID,
    MMDs_start = MMDs,
    HIT6_start = HIT6
  )

cat("Patients with start-of-cycle data:", nrow(c1_start), "\n")


# ------------------------------------------------------------------
# Define end of cycle
# ------------------------------------------------------------------
# End-of-cycle is defined as:
# - Month 12 if available
# - Otherwise, the last observed month within Cycle 1

c1_end <- c1 %>%
  group_by(SUBJECT_ID) %>%
  filter(month_num == max(month_num, na.rm = TRUE)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  select(
    SUBJECT_ID,
    end_month = month_num,
    MMDs_end = MMDs,
    HIT6_end = HIT6
  )

cat("Patients with end-of-cycle data:", nrow(c1_end), "\n")
cat("End month distribution:\n")
print(table(c1_end$end_month, useNA = "ifany"))


# ------------------------------------------------------------------
# Merge start and end and compute changes
# ------------------------------------------------------------------
# Only patients with both start and end are retained

c1_change <- c1_start %>%
  inner_join(c1_end, by = "SUBJECT_ID") %>%
  mutate(
    # Objective change: percentage change in MMDs
    # Negative values indicate improvement
    pct_change_MMDs = (MMDs_end - MMDs_start) / MMDs_start,

    # Subjective change: absolute change in HIT-6
    # Negative values indicate improvement
    delta_HIT6 = HIT6_end - HIT6_start
  )

cat("Patients with valid start/end in Cycle 1:", nrow(c1_change), "\n")

```
The resulting dataset (c1_change) contains one row per subject with Cycle 1 start and end measurements and the corresponding change metrics.
Only subjects with both start-of-cycle and end-of-cycle information are retained (inner_join), ensuring that responder classification is based on complete paired measurements.
===============================================================================
In this next step, we will apply pre-specified clinical thresholds to pct_change_MMDs and delta_HIT6 to define objective and subjective responders and identify discordant profiles. 

Now we classify Cycle 1 responses using clinically motivated thresholds. 
An objective responder is defined as a ≥ 50% reduction in MMDs from Month 1 to Month 12 (pct_change_MMD ≤ -0.50).
A subjective responder is defined as a ≥ 5-point reduction in HIT-6 (delta_HIT6 ≤ -5), corresponding to the Minimal Important Clinical Difference (MICD) reported in the literature.

We then define discordance as objective response without subjective improvement (obj-resp = 1 & subj_resp = 0) and label concordant responders (obj_resp = 1 & subj_resp = 1)
```{r task1-responders-discordance, message=FALSE, warning=FALSE} 
# ------------------------------------------------------------------
# Define responders and discordance (CYCLE 1)
# ------------------------------------------------------------------
# Objective responder:
# - At least 50% reduction in MMDs
#
# Subjective responder:
# - At least 5-point reduction in HIT-6

c1_change <- c1_change %>%
  mutate(
    obj_resp = if_else(
      !is.na(pct_change_MMDs) & pct_change_MMDs <= -0.50,
      1L, 0L
    ),
    subj_resp = if_else(
      !is.na(delta_HIT6) & delta_HIT6 <= -5,
      1L, 0L
    )
  )

cat("Objective responder rate:", mean(c1_change$obj_resp), "\n")
cat("Subjective responder rate:", mean(c1_change$subj_resp), "\n")


# ------------------------------------------------------------------
# Define discordance
# ------------------------------------------------------------------
# Discordance of interest:
# - Objective responder == 1
# - Subjective responder == 0

c1_change <- c1_change %>%
  mutate(
    discordance = if_else(obj_resp == 1L & subj_resp == 0L, 1L, 0L),
    concordant = if_else(obj_resp == 1L & subj_resp == 1L,1L, 0L)
  )

cat("Discordance rate (obj yes, subj no):", mean(c1_change$discordance), "\n")
cat("Concordant responder rate (obj yes, subj yes):", mean(c1_change$concordant), "\n")

# ------------------------------------------------------------------
# Save dataset for downstream analyses (Task 2)
# ------------------------------------------------------------------
dir.create("outputs/task1", recursive = TRUE, showWarnings = FALSE)
readr::write_csv(
  c1_change,
  "outputs/task1/cycle1_responder_discordance_main.csv"
)

dplyr::glimpse(c1_change)
head(c1_change)
```
Using these clinically grounded definitions, we classified each patient at the end of Cycle 1 as an objective responder, a subjective responder, both, or neither.

This classification allows us to explicitly distinguish between:
-**concordant responder**, who show both a clinically meaningful reduction in Monthly Migraine Days and a relevant improvement in patient-reported impact (HI-6);
-**discordant responders**, who achieve a substantial objective improvement (≥50% reduction in MMDs) without a corresponding subjective benefit.

The resulting dataset('c1_change) contains one rwo per subject with Cycle 1 start/end measurements, response indicators and a binary discordance label.
This dataset represents the core input for downstream analyses aimed at  characterizing discordant patients and identifying baseline or longitudinal factors associated wiht objective-subjective mismtach (Task 2).
```{r}
# ------------------------------------------------------------------
# 5.1 2x2 responder table (Objective vs Subjective)
# ------------------------------------------------------------------

tab_2x2 <- c1_change %>%
  count(obj_resp, subj_resp, name = "n") %>%
  mutate(pct_total = n / sum(n))

# Print table
tab_2x2

# Save table for reporting
readr::write_csv(tab_2x2, "outputs/task1_responder_2x2.csv")


# Optional: a more readable labeled table
tab_2x2_labeled <- c1_change %>%
  mutate(
    obj_resp_lbl = if_else(obj_resp == 1L, "Objective responder", "Objective non-responder"),
    subj_resp_lbl = if_else(subj_resp == 1L, "Subjective responder", "Subjective non-responder")
  ) %>%
  count(obj_resp_lbl, subj_resp_lbl, name = "n") %>%
  mutate(pct_total = n / sum(n))

tab_2x2_labeled
readr::write_csv(tab_2x2_labeled, "outputs/task1_responder_2x2_labeled.csv")


# ------------------------------------------------------------------
# 5.2 Scatter plot: objective vs subjective change (Δ vs Δ)
# ------------------------------------------------------------------
# X axis: % change in MMDs (end vs start)
# Y axis: change in HIT-6 (end - start)
# Thresholds:
# - Objective responder: pct_change_MMDs <= -0.50
# - Subjective responder: delta_HIT6 <= -5

# Create a quadrant label for interpretation (optional but useful)
c1_change <- c1_change %>%
  mutate(
    quadrant = case_when(
      obj_resp == 1L & subj_resp == 1L ~ "Concordant responders",
      obj_resp == 1L & subj_resp == 0L ~ "Discordant (obj yes, subj no)",
      obj_resp == 0L & subj_resp == 1L ~ "Discordant (obj no, subj yes)",
      TRUE ~ "Concordant non-responders"
    )
  )

# Scatter plot
p_scatter <- ggplot(c1_change, aes(x = pct_change_MMDs, y = delta_HIT6)) +
  # Decision thresholds
  geom_vline(xintercept = -0.50, linetype = "dashed") +
  geom_hline(yintercept = -5, linetype = "dashed") +
  # Points (shape indicates discordance)
  geom_point(aes(shape = factor(discordance)), alpha = 0.85) +
  labs(
    title = "Objective vs Subjective Change (Cycle 1)",
    x = "MMDs % change (end vs start)",
    y = "HIT-6 change (end - start)",
    shape = "Discordance (1 = obj yes, subj no)"
  )

p_scatter

# Save the figure
ggsave("outputs/task1_scatter_delta_vs_delta.png", p_scatter, width = 7, height = 5)


# ------------------------------------------------------------------
# 5.3 Example counts per quadrant (quick summary)
# ------------------------------------------------------------------

quad_summary <- c1_change %>%
  count(quadrant, name = "n") %>%
  mutate(pct = n / sum(n)) %>%
  arrange(desc(n))

quad_summary
readr::write_csv(quad_summary, "outputs/task1_quadrant_summary.csv")
```

